Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-07-03 CGP Journal/main.tex
Most of the updates suggested in our meeting, still need to move semantic section and reference Equation 1 in discussion section

\section{Genome Ordering}
\label{sec:ordering}
As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
to prevent cycles, by ensuring nodes only receive input from sources that
precede them in the genome.  This restriction does not limit CGP's ability
to represent DAGs, as all DAGs can be serialized to fit this requirement.
Yet this likely has an impact on CGP's ability to evolve specific
DAGs~\cite{goldman:2013:ordering}.

@@

@@

\section{Qualitative Comparisons}
\label{sec:quality}
Empirical results are an important step in confirming our examinations.

We accomplish this by testing the effectiveness
of the described variants and their combination in problem solving.  Deeper analysis of
evolutionary side effects are given in Section~\ref{sec:analysis}.

@@

To prevent undue biasing in the final results, these tuning runs were
not used in the subsequent data analysis.  Effectively, the runs used to set
the parameter configuration were considered
the training runs for that configuration.  As a result, our final results were gathered
from a completely independent set of runs to test the configuration's general characteristics.
Each configuration was run 50 times to ensure statistical power.

@@

@@

\subsection{Scalability}
TODO Discuss using tuned parameters on different problem sizes and analyze
results.  Include O(N) table similar to previous paper.

TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to an active node.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.

>>>>

\section{Genome Ordering}
\label{sec:ordering}
As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
to prevent cycles, by ensuring nodes only receive input from sources that
precede them in the genome.  This restriction does not limit CGP's ability
to represent DAGs, as all DAGs can be serialized to fit this requirement.
However, this representation likely has an impact on CGP's ability to evolve specific
DAGs~\cite{goldman:2013:ordering}.

@@

As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
derived the formula which predicts the probability a node is active based on its
index, given in Equation~\ref{eq:prob} and shown in Figure~\ref{fig:prob}.
\begin{equation}
p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^a)
\label{eq:prob}
\end{equation}
In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
$O$ is the number of output locations, and $a$ is the arity of each node.
This equation is constructed as a negation, such that we determine the probability
the node is not active due to each potential source that could make it active.
The product term calculates the probability that the node at index $j$ is connected to the
node at index $i$, times the probability that the node at index $j$ is active.  This is
negated to determine the probability node $i$ is not active because of node $j$.
All together, the product term calculates the probability that no active node is
connected to the node at index $i$.  Nodes can also become active if they are directly
connected to by an output location.  As such the product is scaled by the probability
that all output locations connect to inputs or nodes not at index $i$.  This
formula assumes only operator bias, not selective pressure, is acting on the
genome.

\begin{figure}
\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
         at a given index will be active.  $N=2000, a=2, I=6, O=6$.}
\label{fig:prob}
\end{figure}

From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
are almost certainly going to be active, while nodes in the last three quarters
of the genome are almost never going to be active.

@@

\section{Qualitative Comparisons}
\label{sec:quality}

\begin{table}
	\centering
	\begin{tabular}{|c|p{6cm}|}
	\hline
  \multicolumn{2}{|c|}{\textbf{Duplication}} \\ \hline
	\textbf{\emph{Skip}} & Whenever an actively identical offspring is detected, assign it its parent's fitness. \\ & \\
	\textbf{\emph{Accumulate}} & Repeatedly perform full genome mutation until \textbf{one or more} active genes change. \\ & \\
	\textbf{\emph{Single}} & Select and mutate individual genes until \textbf{exactly one} active gene changes. \\
	\hline \multicolumn{2}{|c|}{\textbf{Ordering}} \\ \hline
	\textbf{\emph{Normal}} & Nodes must take input from other nodes or inputs that precede them in the genome. \\ & \\
	\textbf{\emph{Reorder}} & Each generation the parent's node ordering is shuffled without effecting node behavior. \\ & \\
	\textbf{\emph{DAG}} & Connection genes can mutate to make any connection as long as no cycle is formed. \\
	  \hline
	\end{tabular}
	\caption{Short descriptions of each of the algorithm variants.}
	\label{tab:glossary}
\end{table}

Empirical results are an important step in confirming our examinations.
We accomplish this by testing the effectiveness
of the described variants and their combination in problem solving.  Deeper analysis of
evolutionary side effects are given in Section~\ref{sec:analysis}.
Table~\ref{tab:glossary} gives a summary of what each variant means.

@@

To prevent undue biasing in the final results, these tuning runs were
not used in the subsequent data analysis.  Effectively, the runs used to set
the parameter configuration were considered
the training runs for that configuration.  As a result, our final results were gathered
from a completely independent set of runs to test the configuration's general characteristics.
Each configuration was run 50 times to ensure statistical power.  All told over
11.5 billion evaluations were performed to gather our results.

@@

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP and CGP using modules in~\cite{walker:2008:cgpmodules}
for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
The same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
CGP with modules required over 800 thousand evaluations, worse than our \emph{Skip-Normal},
and approximately 4 times more than \emph{Skip-Reorder}.  We must be wary about comparisons
to their 3-bit multiplier though, as their version used a different function set.
That said the function set they used appears to have been chosen explicitly to improve
CGP's performance on that problem, so we suggest that if we had used a the same function
set our results would still be significantly better.

@@

\begin{comment}
\subsection{Scalability}
TODO Discuss using tuned parameters on different problem sizes and analyze
results.  Include O(N) table similar to previous paper.

TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.
\end{comment}

@@

Each diagram describes node behavior as falling into eight distinct categories
split between active and inactive nodes.  Nodes are put into the first category
they satisfy.
\begin{enumerate}
\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
\item ``Intron'' - Any node with semantics identical to a node that was
                   active before applying \Call{Simplify}{}, but was inactive afterward.
\item ``Explore'' - All remaining nodes.
\end{enumerate}
Used nodes are by definition only in the active portion of the genome.  Similarly
Explore nodes can only be in the inactive portion of the genome.   Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.
